---
title: "Data Exploration with the Titanic Dataset"
author: "Pier Lorenzo Paracchini, `r format(Sys.time(), '%d.%m.%Y')`"
output: 
  html_document: 
    keep_md: yes
---

```{r echo = FALSE, results='hide'}
Sys.setlocale("LC_ALL", "C")
```

Data available on [Kaggle Titanic Competition](https://www.kaggle.com/c/titanic)  

## Some more information  
_'The sinking of the RMS Titanic is one of the most infamous shipwrecks in history.  On April 15, 1912, during her maiden voyage, the Titanic sank after colliding with an iceberg, killing 1502 out of 2224 passengers and crew. This sensational tragedy shocked the international community and led to better safety regulations for ships. One of the reasons that the shipwreck led to such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others, such as women, children, and the upper-class.In this challenge, we ask you to complete the analysis of what sorts of people were likely to survive. In particular, we ask you to apply the tools of machine learning to predict which passengers survived the tragedy.'_

## Data Exploration

### Data Description

    VARIABLE DESCRIPTIONS:
    survival        Survival
                    (0 = No; 1 = Yes)
    pclass          Passenger Class
                    (1 = 1st; 2 = 2nd; 3 = 3rd)
    name            Name
    sex             Sex
    age             Age
    sibsp           Number of Siblings/Spouses Aboard
    parch           Number of Parents/Children Aboard
    ticket          Ticket Number
    fare            Passenger Fare
    cabin           Cabin
    embarked        Port of Embarkation
                    (C = Cherbourg; Q = Queenstown; S = Southampton)
    
    SPECIAL NOTES:
    Pclass is a proxy for socio-economic status (SES)
     1st ~ Upper; 2nd ~ Middle; 3rd ~ Lower
    
    Age is in Years; Fractional if Age less than One (1)
     If the Age is Estimated, it is in the form xx.5
    
    With respect to the family relation variables (i.e. sibsp and parch)
    some relations were ignored.  The following are the definitions used
    for sibsp and parch.
    
    Sibling:  Brother, Sister, Stepbrother, or Stepsister of Passenger Aboard Titanic
    Spouse:   Husband or Wife of Passenger Aboard Titanic (Mistresses and Fiances Ignored)
    Parent:   Mother or Father of Passenger Aboard Titanic
    Child:    Son, Daughter, Stepson, or Stepdaughter of Passenger Aboard Titanic
    
    Other family relatives excluded from this study include cousins,
    nephews/nieces, aunts/uncles, and in-laws.  Some children travelled
    only with a nanny, therefore parch=0 for them.  As well, some
    travelled with very close friends or neighbors in a village, however,
    the definitions do not support such relations.

### Loading the data

When loading the data 2 consideration 

- avoid to use string as factors (e.g with the `Name` and `Tickect` features) and 
- use the empty field as `NA`.

```{r loadTheData, echo = TRUE}
dataFile <- "./train.csv"
train.data <- read.csv(dataFile, header = TRUE, na.strings = c("NA", ""), stringsAsFactors = FALSE)
rawData <- train.data
```

Dataset structure - we can see that some of the vailable features can be transformed into categorical data.

```{r datasetStructure}
str(rawData)
```

### Preparing the data

```{r commonFunctions, echo = T}
factorize <- function(x, y){
    if (any(is.na(x))) {
        x <- factor(x, exclude = NULL, labels = c(y, "NA"))
    } else {
        x <- factor(x, exclude = NULL, labels = y)
    }
    x
}
```


#### Transforming categorical data into `factors`  

Some of the features are defining categorical data like `Survived, Pclass, Sex, Embarked`. Let's transform them into factors..  
Note that when transforming into a categorical variable __NAs are preserved__.  

```{r transformingSurvived, collapse=TRUE}
any(is.na(rawData$Survived))
##factorized function manages NAs
rawData$Survived_f <- factorize(rawData$Survived, c("No", "Yes"))
str(rawData$Survived_f)
table(rawData$Survived_f)
```

```{r transformingPclass, collapse=TRUE}
any(is.na(rawData$Pclass))
##factorized function manages NAs
rawData$Pclass_f <- factorize(rawData$Pclass, c("1st", "2nd", "3rd"))
str(rawData$Pclass_f)
table(rawData$Pclass_f)
```

```{r transformingSex, collapse=TRUE}
any(is.na(rawData$Sex))
##factorized function manages NAs
rawData$Sex_f <- factorize(rawData$Sex, c("F", "M"))
str(rawData$Sex_f)
table(rawData$Sex_f)
```

```{r transformingEmbarqued, collapse=TRUE}
any(is.na(rawData$Embarked))
##factorized function manages NAs
rawData$Embarked_f <- factorize(rawData$Embarked, c("C", "Q", "S"))
str(rawData$Embarked_f)
table(rawData$Embarked_f)
```

Transformed data now looks like this...

```{r transfDatasetStructure}
#Note at the new features that have been added to the original dataset
str(rawData)
```

#### Finding Missing Values  

Checking if features containg `NA`s and `NaN`s ..

```{r findMissingValues}
countNAs <- function(x){
    if(is.factor(x)){
        sum(x == "NA")
    }else{
        sum(is.na(x))    
    }
}

countNaNs <- function(x){
    sum(is.nan(x))
}

NaN_info <- sapply(rawData, countNaNs)
NaN_info

NA_info <- sapply(rawData, countNAs)
NA_info
```

From tha output we can see that there are not `NaN`s in the dataset, but we have some `NA`s  

- `Age` -> `r sum(is.na(rawData$Age))` (`r round(100* sum(is.na(rawData$Age))/ length(rawData$Age),2)`%)
- `Cabin` -> `r sum(is.na(rawData$Cabin))` (`r round(100* sum(is.na(rawData$Cabin))/ length(rawData$Cabin),2)`%)
- `Embarked` -> `r sum(is.na(rawData$Embarked))` (`r round(100* sum(is.na(rawData$Embarked))/ length(rawData$Embarked),2)`%)

Percentage of `NA`s

```{r percentageMissingValues}
percNAs <- function(x){
    if(is.factor(x)){
        round(100* sum(x == "NA")/ length(x),3)
    }else{
        round(100 * sum(is.na(x))/ length(x),3)
    }
}
sapply(rawData, percNAs)
```

#### Managing missing values: the strategy

`Cabin` contains a high number of `NA`s (around 70%) so it is quite difficult to find a replecement strategy. Another option could be to remove the obeservations with NAs but we will lose around 70% of the observation.

```{r removingCabin}
rawData$Cabin <- NULL
```

`Embarked` contains a very limited number of `NA`s so the strategy that could be used for the missing values is to use the more probable setting for that specific feature. Another possible option could be to remove such observations with the missing value.

```{r managingEmbarked, collapse = TRUE}
table(rawData$Embarked, useNA = "always")
#Note that S is the more probable port so we can set the missing values to S
indexEmbarked_NAs <- which(is.na(rawData$Embarked))
rawData$Embarked[indexEmbarked_NAs] <- "S"
rawData$Embarked_f[indexEmbarked_NAs] <- "S"
table(rawData$Embarked, useNA = "always")
table(rawData$Embarked_f, useNA = "always")
```

`Age` contains around 20% of observations with a missing value.
One possible strategy is to identify the connection between the title and age for the observation with `age` value. Use then title to provide the best guess of the age (for observatios with missing age).

```{r showAgeData}
head(rawData[, c("Name", "Age")], 20)
```

```{r managingAge}
words_tbl <- table(unlist(strsplit(rawData$Name, "\\s+")))
sort(words_tbl[grep("\\.", names(words_tbl))], decreasing = TRUE)

#Knowing the title associated to a missing value age can give some informtion oh how
#to recover the missing age
indexAge_NAs <- which(is.na(rawData$Age))
indexAge_notNAs <- which(!is.na(rawData$Age))

#What is the suvival status for passengers with a missing age
table(rawData$Survived_f[indexAge_NAs])

passengersTitle <- rawData[, c("Age", "Name")]
passengersTitle$StartPos <- regexpr("[a-zA-Z]+\\.", passengersTitle$Name)
passengersTitle$EndPos <- regexpr("\\.\\s+[a-zA-Z(]", passengersTitle$Name)
passengersTitle$Title <- substr(passengersTitle$Name, passengersTitle$StartPos, passengersTitle$EndPos)
```


Title in the observations with calculate average age for title (excluding observation having age value set to `NA`).

```{r titleAgeInfoSummary}
table(passengersTitle$Title[indexAge_notNAs])
titleMeans <- aggregate(passengersTitle$Age[indexAge_notNAs], by = list(title = passengersTitle$Title[indexAge_notNAs]), mean)
titleMeans
```

Adding ages for the observation with missing values `NA`s...

```{r ageNAsReplacement}
gestBestGuessOnAge <- function(title){
    ##titleMeans is a variable defined in the global environment
    titleMeans$x[titleMeans$title == title]
}

guessedAge <- sapply(passengersTitle$Title[indexAge_NAs],FUN = gestBestGuessOnAge)
rawData$Age[indexAge_NAs] <- round(guessedAge)
head(rawData[, c("Name", "Age")], 20)
```
### Visualizing the Data

```{R visualizations}
barplot(table(rawData$Survived), main = "Passenger Survivals", names = c("Died", "Survived"))
barplot(table(rawData$Pclass_f), main = "Passenger Class")
barplot(table(rawData$Sex), main = "Passenger Gender")
hist(rawData$Age, main = "Passenger Age Distribution",breaks = 20)
barplot(table(rawData$SibSp), main = "Number of Siblings OnBoard")
barplot(table(rawData$Parch), main = "Number of Parents/ Children OnBoard")
hist(rawData$Fare, main = "Passenger Fare Distribution")
barplot(table(rawData$Embarked), main = "Port of Origin")
```

```{r someMoreVisualization}
barplot(table(rawData$Survived, rawData$Sex_f), main = "Passenger Survival by Gender", col = c("red", "green"), legend = c("Died", "Survived"))
barplot(table(rawData$Survived, rawData$Pclass_f), main = "Passenger Survival by Class", col = c("red", "green"), legend = c("Died", "Survived"))
barplot(table(rawData$Sex_f, rawData$Pclass_f), main = "Passenger Gender by Class", col = c("pink", "blue"), legend = c("Female", "Male"))

survivalBySexClass <- as.data.frame(table(rawData$Survived, rawData$Pclass_f, rawData$Sex_f))
survivalBySexClass_f <- subset(survivalBySexClass, survivalBySexClass$Var3 == "F")
survivalBySexClass_m <- subset(survivalBySexClass, survivalBySexClass$Var3 == "M")
sf_matrix <- matrix(survivalBySexClass_f$Freq, nrow = 2, ncol = 3)
rownames(sf_matrix) <- c("0", "1")
colnames(sf_matrix) <- c("1st", "2nd", "3rd")
barplot(sf_matrix, col = c("red", "green"), main = "Survival Female Passenger By Class", legend = c("Died", "Survived"))

sf_matrix <- matrix(survivalBySexClass_m$Freq, nrow = 2, ncol = 3)
rownames(sf_matrix) <- c("0", "1")
colnames(sf_matrix) <- c("1st", "2nd", "3rd")
barplot(sf_matrix, col = c("red", "green"), main = "Survival Male Passenger By Class", legend = c("Died", "Survived"))
```

```{r someSomeMoreVisualization}
index_survived <- which(rawData$Survived_f == "Yes")
index_notSurvived <- which(rawData$Survived_f == "No")
hist(rawData$Age[index_notSurvived], main = "Passenger Age Distribution", xlab = "Age", ylab = "No Of Passengers", col = "red", breaks = seq(0,80, by=2))
hist(rawData$Age[index_survived], col = "green", breaks = seq(0,80, by=2), add = TRUE)
legend("topright", c("Died", "Survived"), col=c("red", "green"), lwd=10)
```

```{R andSomeMore}
boxplot(rawData$Age ~ rawData$Survived_f, main = "Passenger by Age", xlab = "Survived", ylab = "Age")
```

```{R categorizedPassengerByAge, collapse = TRUE}
rawData$isChild <- rawData$Age < 13
rawData$isAdolescent <- rawData$Age >= 14 & rawData$Age < 20
rawData$isAdult <- rawData$Age >= 20 & rawData$Age < 65
rawData$isSenior <- rawData$Age >= 65
table(rawData$Survived_f, rawData$isChild)
sum(rawData$Survived_f == "Yes" & rawData$isChild == TRUE)/ sum(rawData$isChild)

table(rawData$Survived_f, rawData$isAdolescent)
sum(rawData$Survived_f == "Yes" & rawData$isAdolescent == TRUE)/ sum(rawData$isAdolescent)

table(rawData$Survived_f, rawData$isAdult)
sum(rawData$Survived_f == "Yes" & rawData$isAdult == TRUE)/ sum(rawData$isAdult)

table(rawData$Survived_f, rawData$isSenior)
sum(rawData$Survived_f == "Yes" & rawData$isSenior == TRUE)/ sum(rawData$isSenior)
```
